---
title: "03-solution"
author: "M Hallett"
date: "22/10/2021"
output:
  html_document:
    df_print: paged
  revealjs::revealjs_presentation:
    css: solar.css
    incremental: yes
    theme: black
    highlight: haddock
    transition: convex
    fig_width: 7
    fig_height: 6
    fig_caption: yes
    self_contained: no
    reveal_plugins:
    - search
    - zoom
    - notes
    - chalkboard
    - menu
    reveal_options:
      slideNumber: yes
      previewLinks: yes
      chalkboard:
        theme: chalkboard
        toggleNotesButton: yes
      menu:
        numbers: yes
---



_Part1 1_


You can do it in R. (I don't want to run this so it's commented out.)
```{r, eval=FALSE}
# Create raw directory
#dir.create("/cloud/project/raw")

# Download file
#download.file(url = "ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/taxdmp.zip", 
#              destfile = "/cloud/project/raw/taxdmp.zip",
#              method = "wget")

# Unzip
#unzip("/cloud/project/raw/taxdmp.zip", exdir = "/cloud/project/raw/taxdmp")
```

Or you can do it in Unix.

```{r, eval=FALSE}
# Create raw directory
#mkdir /cloud/project/raw

# Download file
#wget ftp://ftp.ncbi.nlm.nih.gov/pub/taxonomy/taxdmp.zip \
      /cloud/project/raw/taxdmp.zip

# Unzip
#cd /cloud/project/raw
#gunzip /cloud/project/raw/taxdmp
```

_Part 2_

```{r, eval=FALSE}
require(purrr); require(tidyverse)
```

```{r, eval=FALSE}
# File names + dir
files <- c("division" = "/cloud/project/raw/taxdmp/division.dmp", 
           "nodes" = "/cloud/project/raw/taxdmp/nodes.dmp", 
           "names" = "/cloud/project/raw/taxdmp/names.dmp")

dir.create("/cloud/project/refined")

# A solution from last year's TA Aky Eftyhios (many other ways to do this.)
read_dmp <- function(dmp_file) {
  out <- readLines(dmp_file) %>%
    gsub(pattern = "\t\\|$", replacement = "") %>% 
    str_split(., pattern = "\t\\|\t", simplify = TRUE) %>%
    as_tibble()
  
  return(out)
}

res <- map(.x = files, .f=read_dmp)
```

You'll see that the tables do no thave column names with Aki's solutions. We could carefully add them here. As a side note: this is not my preferred solution; it can be dangerous to lose the original column names rather than changing them individually afterwards. I also don't find optimal the use of spaces in the names or the mixed upper and lower case, as discussed in the lecture.

```{r, eval=FALSE}
fields <- list("division" = c("division id", "division cde", "division name", "comments"),
               
               "nodes" = c("tax_id", "parent tax_id", "rank", "embl code", "division id",
                           "inherited div flag", "genetic code id", "inherited GC flag", 
                           "mitochondrial genetic code id", "inherited MGC flag",
                           "GenBank hidden flag", "hidden subtree root flag", "comments"),
               
               "names" = c("tax_id", "name_txt", "unique name", "name class")
               )

colnames(res$nodes) <- fields$nodes
colnames(res$division) <- fields$division
colnames(res$names) <- fields$names
```

Now we just save the refined version.
```{r, eval=FALSE}
dir.create("/cloud/project/refined")
#saveRDS(res, "/cloud/project/refined/master.rds")
```

_Part 3_

```{r, eval=FALSE}
filter_terms <- c("comments", "mitochondrial genetic code id", "inherited div flag",
                  "genetic code id", "inherited GC flag", "inherited MGC flag", 
                  "GenBank hidden flag", "hidden subtree root flag")

# Filter
refined <- map(
  .x = res,
  .f = ~ select(.x, -any_of(filter_terms))
  )
saveRDS(refined, "/cloud/project/refined/master.rds")  # save updated version
```

_Part 4_ 

Is it tidy? Well... As I mentioned above, I don't like the column names in some cases but that's a minor point. Mostly the three tibbles are tidy, except for ${\tt unique name}$ in my opinion. There is both the name name of the taxon and then extra information in these angle brackets. We'll leave it for now though. 

Do you agree?

_Part 5_ 

```{r, eval=FALSE}
str(refined)
```


No there are many problems here mostly where characers like ${\tt tax\id}$ probably should be integers or numeric.

```{r, eval=FALSE}
refined$division[["division id"]] <- as.numeric(refined$division[["division id"]])

# Nodes
refined$nodes <-  refined$nodes %>%
  mutate_at(vars("tax_id", "parent tax_id", "division id"), ~ as.numeric(.))

# Names
refined$names[["tax_id"]] <- as.numeric(refined$names[["tax_id"]])

# Print
str(refined)
```


_Part 6_

Is ${\tt tax\_id}$ unique in its own table? That is, does each observation have a unique ${\tt tax\_id}$? Yes, it can serve as a primary key in the ${\tt refined\$nodes}$ tibble.
```{r, eval=FALSE}
refined$nodes %>%
  count(tax_id) %>% 
  filter(n > 1) %>% 
  nrow
```

It is also present in the ${\tt names}$ tibble but it i not unique, so it serves as the foreign key.

```{r, eval=FALSE}
refined$names %>%
  count(tax_id) %>% 
  filter(n > 1) %>%
  nrow
```

We can check if any observation in either table is NA.

```{r, eval=FALSE}
length(which(is.na(refined$nodes["tax_id"]))); length(which(is.na(refined$names["tax_id"])))

```

We can use ${\tt left\_join$  because we want to keep all the observations in 
${\tt refined\$names}$. These entries match ${\tt refined\$nodes}$, therefore we don't lose any data. It keeps the one-to-many relationship intact

```{r, eval=FALSE}
tbl <- refined$names %>%
  left_join(refined$nodes, by = "tax_id")
tbl
```

_Part 7_

Primary key is ${\tt division id}$ because it uniquely 
identifies each observation. 

```{r, eval=FALSE}
refined$division %>%
  count(`division id`) %>% 
  filter(n > 1) %>%
  nrow
```

Foreign key is `division id` in our new tbl from Part 6 because it appears in the
refined$division table, where it matches each division to a unique id.

We use the ${\tt left\_join}$ because we would like to keep all the observations in the original tibble (${\tt tbl}$) and add the additional information found in the ${\tt dividsion}$ tibble.


```{r, eval=FALSE}
tbl <- tbl %>%
    left_join(refined$division, by = "division id")
```

_Part 8_

```{r, eval=FALSE}
get_taxon_name <- function(tbl, target_tax_name) {
  tmp <- tbl %>% filter(name_txt == target_tax_name) %>% .[["tax_id"]]
  return(tmp)
}
```

_Part 9_

```{r, eval=FALSE}
get_children <- function(tbl, target_tax_id) {
  tmp <- tbl %>% filter(`parent tax_id` == target_tax_id) 
  return(tmp)
}
```

_Part 10_

This is a very hard question. First we write a function to return all of the species below a node in the tree

```{r, eval=FALSE}
collect_all <- function(tbl, start_tax_id, target_rank="species") {
  
  to_do <- start_tax_id  # this vector keeps track of every taxon we need to visit
  good_ones <- c()  # use this to store all tax_id that have the right rank
  
  while (length(to_do)>0) {   # we are finished when it's empty
    
    current <- to_do[1]   # look at the first entry of to_do
    to_do <- to_do[-1]  # remove first entry from to_do
    
    tmp <- tbl %>% filter(tax_id==current) %>% .[['rank']]
    
    # check if one of them is the right rank (there could be more than one)
    for (j in 1:length(tmp)) {
      if (tmp[j]==target_rank) {
        good_ones <- c(good_ones, current)
        break
      }
    }
    currents_kids <- unique(get_children(tbl, current)[['tax_id']])
    to_do <- c(to_do, currents_kids) 
  }
  return(length(good_ones))
}
```

Ok so now all that is left is to run this for the three kingdoms first with rank species and then rank genera.

```{r, eval=FALSE}
bacteria_tax_id <- 2
eukaryota_tax_id <- 2759
archaea_tax_id <- 2157
# 
# bac_num_species <- collect_all(tbl, start_tax_id=bacteria_tax_id, target_rank="species") 
# (bac_num_species)
# euk_num_species <- collect_all(tbl, start_tax_id=eukaryota_tax_id, target_rank="species") 
# (euk_num_species)
# arch_num_species <- collect_all(tbl, start_tax_id=archaea_tax_id, target_rank="species") 
# (arch_num_species)
# 
# bac_num_genera <- collect_all(tbl, start_tax_id=bacteria_tax_id, target_rank="genus") 
# (bac_num_genera)
# euk_num_genera <- collect_all(tbl, start_tax_id=eukaryota_tax_id, target_rank="genus") 
# (euk_num_genera)
# arch_num_genera <- collect_all(tbl, start_tax_id=archaea_tax_id, target_rank="genus") 
# (arch_num_genera)
```


_Part 11_

Let's do Part _11_ first. Afterwards _Part 10_ is easier. Here our goal is to 
write a function that returns all the taxa below a given _division_ in the tree. There
are many ways to do this.

```{r, eval=FALSE}
get_all_division_taxa <- function(tbl, division) {
  nodes <- tbl %>% filter(`division id`==division)
  return(nodes)
}
```


_Part 12_

Here is a solution from a previous year (Aki Eftyhios). I provide a recursive solution below (beyond the scope of the course).

```{r, eval=FALSE}
path_to_root <- function(tax_id, taxonomy) {
  
  # Function does what the question asks :) 
  
  # Using the environment because of undefined behavior when trying to filter with
  # tax_id. The names conflict.
  functione_env <- environment()
  path <- c()
  
  # Obvious, but still need to instantiate. 
  # Hard coding it could be a problem if they ever change the root tax_id, 
  # but I doubt that will ever happen.
  root_tax_id <- 1
  
  # Functions as off switch in while loop. Allows for one iteration using root
  count_switch <- 2
  
  while (count_switch != 0) {
    
    # Get name of node and add to path vector
    node_name <- taxonomy %>%
      filter(`tax_id` == get("tax_id", functione_env)) %>%
      pull("name_txt") %>%  # Pull name
      replace_na("NA") # Fixes NAs
    
    # Add to vector
    path <- c(path, node_name) 
    
    # Switch to new node
    tax_id <- taxonomy %>%
      filter(`tax_id` == get("tax_id", functione_env)) %>%
      pull("parent tax_id")
    
    # Off switch
    if (tax_id == root_tax_id) {
      count_switch <- count_switch - 1
    }
    
  }
  return(cat(rev(path), sep = ", "))
}
```

Here is a recursive solution (as discussed in class; beyond the scope of the course).

```{r, eval=FALSE}

path_to_root <-  function(tbl, current_tax_id, target_tax_id) {

  tmp <- tbl %>% filter(tax_id ==  current_tax_id)
  if (nrow(tmp)>1) tmp <- tmp[1,]  # annoying if multiple matches, so just take the first one.
  
  if (current_tax_id == target_tax_id) {
    return(tmp[['name_txt']])
  } 
  parent <- tmp %>% .[['parent tax_id']]
  return(c(tmp[['name_txt']], path_to_root(tbl, parent, target_tax_id  )))
}

# for some reason the NCBI set the parent of the root to be the parent.
# that doesn't make any sense. Fix here.
fix <- which(tbl$tax_id == 1)
tbl[fix, 'parent tax_id'] <- NA

x <- path_to_root(tbl, current_tax_id = 2157, target_tax_id=1)
# path from archae to root
print(x)
```




